---
title: "Soil Water Evaporation"
author: "Rachel Havranek"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document: 
    df_print: paged
    toc: yes
    234toc_depth: 2
    toc_float: true
    code_folding: show
---


```{r "load packages", include=FALSE, message=FALSE}
library(tidyverse)
```

# Introduction
This model closely follows Allison and Barnes, 1983 & Barnes and Allison, 1988. This model has three parts:\
1. Exploring the isotope profile of an isothermal saturated soil\
2. Exploring the isotope of an isothermal unsaturated soil\
3. Exploring the isotope profile of a summertime unsaturated soil

# Saturated Soil 
The change in the isotope ratio with respect to depth of a saturated soil can be described by the equation(After Zimmermann et al., 1967:):
*This assumes that pore space is constant through the soil column.*
$$
\frac{dR}{dz} = \frac{E(R-R_{res})}{D^*} \\
where:\\
E = \text{Evaporation rate } [m/sec]\\
p = total\ porosity\ \\
\tau = tortuosity \\
D_l = \text{diffusion coefficient of liquid water } [m^2/sec]\\
D^* = \text{effective diffusivity of the isotope in the pore water} = (p\tau D_l)\\
$$

The solution of the equation above is:
$$
R = R_{res} + (R_0 - R_{res})*e^{(-z/zl)} \\
where: \\
z_l = D^*/E
$$

We can express the above equation in delta notation:  
$$
\delta  = \delta_{res} +(\delta_{surface} - \delta_{res})*e^{(-z/z_l)} \\
$$

### Activity 1: In breakout rooms, describe in words what's going on in this equation (pairs). Can you relate that back to the craig-gordon model of evaporation?

Okay! So let's make a plot of what an isotope profile of a saturated soil might look like: 
```{r}
saturated_soil <- tibble(
  # set up 
  z = seq(0,.25, by = .001), #m
  
  #Oxygen isotope boundary conditions
  d18O_res = -4, #permil
  d18O_surface = 0, #permil
  
  #Soil boundary conditions
  E = 10E-9, #m/sec - this is just under 1 mm / day (0.864 mm/ day)
  p = 0.3, # porosity, unitless
  tau = 0.67, #toruosity, unitless
  theta = 0.3, #volumetric water content unitless
  D = 1.5E-9, #m^2 / sec from Barnes and Allison, 1988
  f = tau*theta,

  
  #calculated terms
  D_star = f*D,  #m^2 / sec
  z_l = D_star/E, #m
  d18O_z = d18O_res + (d18O_surface - d18O_res)*exp(-z/z_l) #permil
)

saturated_soil %>% ggplot()+aes(x = d18O_z, y = z) + geom_line()  + scale_y_reverse() + scale_x_continuous(limits = c(-4,0)) + theme_bw()
```

### Activity 2: Picarro water isotope analyzers (like those you might find in IsoLab in SEEC) have a precision of usually +/- 0.25 permil in oxygen isotope space. To what depth would you be able to differentiate water that has been partially evaporated from the rest of the soil water? 


# Unsaturated Soil:

#### Water concentration curve:
Let's inspect a soil, with a water concentration profile that looks like this: 
```{r, include = FALSE}
water_C <- tibble (
  z = seq (0,1, by = 0.001),
  theta = ifelse(z<=0.05, 0.05, (z+0.9)/19)
)
  
water_C %>% ggplot () + aes(x = theta, y = z) %>% geom_line()+scale_y_reverse()+scale_x_continuous(limits = c(0,.1)) + theme_bw()
```

**This profile requires that we inspect two parts, the vapor transport zone and the liquid transport zone**


## vapor transport region 
The relative humidity above the evaporitng front in the soil varies linearly with depth 

$$
\frac{dN}{dz} = \frac{\rho E}{D^*_{vapor}} \\
where: \\
\text {N = concentration of water vapor in the soil}\\
\text {rho = density of liquid water}\\
D^*_{vapor} = \text{effective diffusivity of water vapor} = \tau*D^v(\rho-\theta)
$$

so ... 
$$
\frac {dN}{dz} = \frac{\rho E}{\tau*D^v(\rho-\theta)}
$$




We then need to evaluate the relative humidty at each depth. RH at depth is a function of the RH at the surface and a depth scaling factor 
$$
h = h_{atm} + z/\bar{z} \\
$$
z-bar is a function of the saturation, diffusivity of vapor, the density of water, and the evaporative rate.

$$
\bar{z} = \frac{N_{saturation}D^*_{vapor}}{\rho E}\\
$$

The isotope ratio of the liquid at depth can be described by the equation:
$$
R_l = \frac{\alpha_k*R_{reservoir}*z/\bar{z} + h_aR^a}{\alpha_{eq}h}
$$

If we remember that epsilon = alpha - 1, then: 
$$
\delta = \frac{\delta^a +\epsilon_{eq} +(\eta(1+\delta_{res}) + (\delta_{res} - \delta^a)) (z/(z+h_a\bar{z}  ))} {\alpha_{eq}}
$$
hoooo! That's a lot to look at, let's break that up:\

1. a description of how the 'reservoir' and air diffuse:remember, } 
$\eta_i$ = a measure of the ability of molection of species i to escape from the surface
$$
diffusion\ term = \eta(1+\delta_{reservoir}) + (\delta_{reservoir} - \delta^a)\\
and:\\
\eta = (1-h_a)/\epsilon_k
$$
so, the equation breaks up into, how effective are molevules of species i leaving the reservoir, and how differeent are the reservoir and air 


1. a description of what the humidity is at every z:
$$
humidity\ term = \frac{z}{z+h_a\bar{z}}
$$
So, the equation above can be rewritten as: 
$$
\delta = \frac{\delta^a +\epsilon_{eq} +(diffusion)*(humidity)} {\alpha_{eq}}
$$

okay, let's do some math: 
```{r}
vt_region <-  tibble(
  #the basics: 
  z = seq(0,.05, by = .001), #m
  temp = 20, #°C
  temp.K = temp + 273.15,
  rho = 997, #kg/m^3
  
  #fractionation factors:
  R = 8.314, #J/mol-K
  epsilon_d18O_eq = -7.685+6.7123*(1E3/temp.K)-1.6664*(1E6/temp.K^2)+0.35041*(1E9/temp.K^3),
  alpha_eq = exp(epsilon_d18O_eq/1000),
  alpha_kin = .975, 
  epsilon_kin = -25, #permil
  
  #isotope boundary conditions: 
  d18O_res = -4, #permil
  d180_surface = 0, #permil
  d18O_a = -14, #permil
  
  #Soil boundary conditions
  h_a = 0.2, #relative humidity @ surface
  E = 10E-9, #m/sec
  p = 0.3, # porosity, unitless
  tau = 0.67, #toruosity, unitless
  theta = ifelse(z<=0.05, 0.05, (z+0.9)/19), #changes through the profile 
  N_sat = 0.017, #kg/m^2
  D = 1.5E-9, #m^2 / sec from Barnes and Allison, 1988
  D_v = 2.42E-5,
  f = tau*theta,
  D_star = f*D,  #m^2 / sec
  
  #liquid diffusion length scale 
  z_l = D_star/E, #m
  
  
  #kinetic fractionation terms: 
  eta = (1-h_a)/epsilon_kin,
  z_bar = (N_sat*D_v)/(rho*E),
  
  
  diffusion_term  = (eta*(1+d18O_res)) + (d18O_res - d18O_a),
  humidity_term = (z/ (z+h_a*z_bar)),
  d18O =  (d18O_a + epsilon_d18O_eq + diffusion_term*humidity_term)/alpha_eq
  
)
```


## liquid transport zone: 
$$
\delta = \delta_{reservoir} + (\delta_{ef} - \delta_{reservoir})exp({- \int_{zef}^z)\frac{dz}{z_1 +z_v}}\\
z_l = D^*/E \\
z_v = \frac{\alpha_{eq}*\alpha_{kin}*D^*N_{sat}}{E\rho}  
$$
okay, lets do some math. 
```{r}
lt_region <- tibble (
  z = seq (.051, 1, by=0.001),
  z_ef = 0.05, #m
  temp = 20, #°C
  temp.K = temp + 273.15,
  rho = 997, #kg/m^3
  
  #fractionation factors:
  R = 8.314, #J/mol-K
  epsilon_d18O_eq = -7.685+6.7123*(1E3/temp.K)-1.6664*(1E6/temp.K^2)+0.35041*(1E9/temp.K^3),
  alpha_eq = exp(epsilon_d18O_eq/1000),
  alpha_kin = 0.975, 
  epsilon_kin = -25, #permil, average based on Rothfuss et al., 2012
  
  #isotope boundary conditions: 
  d18O_res = -4, #permil
  d18O_surface = 0, #permil
  d18O_a = -14, #permil

  #Soil boundary conditions
  E = 10E-10, #m/sec
  p = 0.3, # porosity, unitless
  tau = 0.67, #toruosity, unitless
  theta = ifelse(z<=0.05, 0.05, (z+0.9)/19), #changes through the profile
  N_sat = 0.017, #kg/m^2
  D = 1.5E-9, #m^2 / sec from Barnes and Allison, 1988
  D_v = 2.42E-5,
  f = tau*theta,
  D_star = f*D,  #m^2 / sec
  g = tau*(1-theta), #this is an approximation (see 1988 paper, p 151)
  Dv_star = g*D_v,
  
  #length scales
  z_l = D_star/E, #m
  z_v = (alpha_eq*alpha_kin*Dv_star*N_sat)/(E*rho),
  k = z_l + z_v, 
  
  #equation 6 from 1988 paper: 
  d18O = d18O_res + (4.35 - d18O_res)*exp(-1/k*(z-z_ef))
)
  
```

```{r}
d18O_profile <- bind_rows(vt_region, lt_region) %>% 
  select(z, d18O) %>% 
    ggplot() + 
    aes(d18O, z)+
    geom_path() + 
    scale_y_reverse()


d18O_profile
```



# Summertime Unsaturated Soil 

so, lets add a more realistic temperature profile: 

  t_ave = 20,
  t_surface = 40
  kappa = 1E-6, # m^2/second, averge value for a loamy soil
  period = 60*60*24*365, #seconds in a year
  z_star  = sqrt((kappa*period)/pi),
  temp = t_ave + (t_surface - t_ave)*exp(-z/z_star)










The isotopic ratio at the top of the saturated surface can be calculated if you know the relative humidity of the atmosphere, the diffusion ratio, the isotopic ratio that is the source, and the isotope composition of the atmosphere
$$
R_{surface}  = \frac{(1-h_a)\sigma R_{res} +h_aR_a)} {\alpha_{eq}} \\
\text{where:}\\
h_a = \text{relative humidity of the atmosphere}\\
\sigma = \text{diffusion ratio} = 1+ (\epsilon_{kinetic}/(1-h_a))  \\
\eta_i = \text{diffusion ratio excess}\\

$$









